### 2013-11-06 src2/thesis.tex
update

@@

This thesis is about applying embedded language techniques to programming 
multi- and many-core computers. Computers today are becoming more and more  
parallel. With each new generation of 
general purpose processor (CPU) or graphics processor (GPU), we get more 
parallel resources. To fully make use of modern CPUs and GPUs, we need to
exploit their parallelism.

While it is sometimes possible to automatically parallelise programs originally 
written for serial execution, it is becoming clear that explicitly parallel 
programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
This also means that it is important to think in parallel from the beginning.

This thesis focuses on data-parallelism, as opposed to control- or 
task-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. Data-parallelism is scalable,
meaning that the potential for parallelism increases with problem size.

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
of cores and number of vector units, varies over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
can use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
its capabilities was also provided. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
towards embedding Intel ArBB in Haskell.

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch prediction and 
ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application.

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
They emphasise the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms. The ArBB library supports flat regular data-parallelism, but also 
irregular and to some degree nested data-parallelism.

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
The programmer needs to implement them using lower level features, making CUDA and OpenCL 
code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
naive reduction pattern is implemented and then refined over many steps, taking many 
characteristics of the GPU into mind. The end result is an implementation of the 
reduction pattern that performs very well on the chosen GPU. But as GPUs evolve, the 
design space may need to be explored again. 
%What is 
%is that says that we do not need to redo this exploration with each new GPU generation 
%and discover a new optimal solution?

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to provide more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian. 
Obsidian is focused on flat and regular data-parallelism.

% \section{The problem}

% \emph{what is so interesting about Parallelism and GPUs}

@@

\begin{center}
{\large \bf This paper will be revised before printing the Thesis} 
\end{center}

>>>>

\usepackage[paperwidth=169mm, paperheight=239mm]{geometry}

@@

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. Embedded languages are languages that are implemented 
as a library for some existing host language. In this case, the functional programming 
language Haskell is used as host language.

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arise. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. {\bf \emph{source??}}

The embedded language approach to programming language implementation is beneficial 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. Other benefits of embedded languages often mentioned are: making use 
of the host language's library functions, if they are general enough. The host language 
can be used as a powerful macro language for the embedded language. An embedded 
language also provide a more leveled learning curve to programmers already familiar 
with the host language. For these programmers using an embedded language is more like  
learning about a new library.

This thesis focuses on data-parallelism, as opposed to task or control parallelism.
Data-parallelism is applicable where the input data can be divided into chunks
and operated upon independently. Data-parallelism is scalable; the potential 
for parallelism increases with the problem size. Two different embedded languages 
for data-parallel programming are presented in this thesis: Obsidian, for 
general purpose computations on GPUs and EmbArBB for parallel programming across 
platforms, CPU, GPU and other accelerators.

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA 
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
of learning about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++.

{\bf {\emph old text from here to end}}

@@

%\begin{center}
%{\large \bf This paper will be revised before printing the Thesis} 
%\end{center}
